---
title: "MATH X314 Assessment 4B Annotated Code"
author: "Pranav Kadari"
format: html
execute:
  echo: true
  warning: false
  message: false
---

# 1. Setup

This file contains the annotated code used to generate the static graphics for the final project portfolio. The comments explain both the transformation steps and the design reasoning behind the graphics.

```{r setup}
library(tidyverse)
library(scales)
library(ggplot2)
```

# 2. Load Data

Place `matches.csv` and `deliveries.csv` inside the `data/` folder before rendering.

```{r load-data}
matches <- read_csv("../data/matches.csv")
deliveries <- read_csv("../data/deliveries.csv")
```

# 3. Shared Theme and Colour Palette

A consistent colour palette is used across the project to address feedback from the previous assessment and create a more coherent visual identity.

```{r theme-palette}
col_primary <- "#1B365D"   # dark blue for core trend/comparison visuals
col_secondary <- "#F28E2B" # orange for distribution visuals
col_accent <- "#B23A48"    # red accent for benchmarks and dense point layers
col_neutral <- "#444444"

base_theme <- theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", colour = col_primary),
    plot.subtitle = element_text(colour = col_neutral),
    plot.caption = element_text(size = 8, colour = col_neutral)
  )
```

# 4. Prepare Match and Innings Data

The uploaded dataset uses `Season` with a capital S. I rename it to `season` so later code remains consistent.

```{r prepare-innings}
matches_clean <- matches %>%
  select(id, Season, team1, team2, winner) %>%
  rename(season = Season) %>%
  mutate(season_year = as.integer(str_extract(season, "\\d{4}"))) %>%
  filter(!is.na(season_year))


# Aggregate ball-by-ball deliveries into innings totals. This creates one row per innings.
innings_totals <- deliveries %>%
  group_by(match_id, inning, batting_team) %>%
  summarise(
    innings_total = sum(total_runs, na.rm = TRUE),
    .groups = "drop"
  )


# First innings are used because they are more comparable across matches than chase innings.
first_innings <- innings_totals %>%
  filter(inning == 1) %>%
  left_join(matches_clean, by = c("match_id" = "id")) %>%
  filter(!is.na(season_year))
```

# 5. Visualisation 1: Average First Innings Score by Season

A line chart is used because the main task is to show change over time.

```{r fig1-average-score-trend}
avg_score_season <- first_innings %>%
  group_by(season_year) %>%
  summarise(
    avg_first_innings_score = mean(innings_total, na.rm = TRUE),
    matches = n(),
    .groups = "drop"
  )

fig1 <- ggplot(avg_score_season, aes(x = season_year, y = avg_first_innings_score)) +
  geom_line(linewidth = 1.1, colour = col_primary) +
  geom_point(size = 2.4, colour = col_primary) +
  labs(
    title = "IPL scoring has risen over time",
    subtitle = "Average first innings score by season",
    x = "Season",
    y = "Average first innings score",
    caption = "Source: Kaggle IPL match and delivery data; first innings only."
  ) +
  scale_x_continuous(breaks = avg_score_season$season_year) +
  base_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

fig1

ggsave("../images/fig1_avg_score_trend.png", fig1, width = 10, height = 6, dpi = 300)
```

# 6. Visualisation 2: Distribution of First Innings Scores

A boxplot is used to show medians, spread, and high-scoring outliers within each season.

```{r fig2-score-distribution}
fig2 <- ggplot(first_innings, aes(x = factor(season_year), y = innings_total)) +
  geom_boxplot(fill = col_secondary, alpha = 0.8, outlier.alpha = 0.55) +
  geom_hline(yintercept = 200, linetype = "dashed", colour = col_accent, linewidth = 0.8) +
  annotate("text", x = 2.2, y = 207, label = "200-run benchmark", colour = col_accent, size = 3.4, hjust = 0) +
  labs(
    title = "High scores are becoming more common",
    subtitle = "First innings score distributions by season",
    x = "Season",
    y = "First innings score",
    caption = "Dashed line marks the 200-run threshold often associated with high-scoring T20 innings."
  ) +
  base_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

fig2

ggsave("../images/fig2_score_distribution.png", fig2, width = 10, height = 6, dpi = 300)
```

# 7. Prepare Batting Data

The uploaded deliveries file uses `batsman`. Wides are excluded from balls faced because they are not legal balls faced by a batter.

```{r prepare-batting}
batting_summary <- deliveries %>%
  group_by(match_id, batsman) %>%
  summarise(
    runs_scored = sum(batsman_runs, na.rm = TRUE),
    balls_faced = sum(wide_runs == 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(balls_faced > 0, runs_scored > 0) %>%
  mutate(strike_rate = (runs_scored / balls_faced) * 100)
```

# 8. Visualisation 3: Runs Scored vs Strike Rate

The scatterplot uses transparency because many batting innings overlap visually.

```{r fig3-runs-strike-rate}
fig3 <- ggplot(batting_summary, aes(x = runs_scored, y = strike_rate)) +
  geom_point(alpha = 0.15, size = 1.2, colour = col_accent) +
  geom_smooth(se = FALSE, linewidth = 0.9, colour = "black") +
  coord_cartesian(ylim = c(0, 450)) +
  labs(
    title = "Successful batting balances speed and volume",
    subtitle = "Runs scored and strike rate for IPL batting innings",
    x = "Runs scored",
    y = "Strike rate",
    caption = "Wides excluded from balls faced; y-axis capped to reduce distortion from extreme short innings."
  ) +
  base_theme

fig3

ggsave("../images/fig3_runs_vs_strike_rate.png", fig3, width = 10, height = 6, dpi = 300)
```

# 9. Prepare Team Win Rate Data

Each match is converted into two team rows so both teams are counted as having played.

```{r prepare-team-win-rate}
team_wins <- matches %>%
  filter(!is.na(winner), winner != "") %>%
  select(id, team1, team2, winner)

team_matches <- bind_rows(
  team_wins %>% transmute(id, team = team1, winner),
  team_wins %>% transmute(id, team = team2, winner)
)

team_summary <- team_matches %>%
  group_by(team) %>%
  summarise(
    matches_played = n(),
    wins = sum(team == winner, na.rm = TRUE),
    win_rate = wins / matches_played,
    .groups = "drop"
  ) %>%
  filter(matches_played >= 10) %>%
  arrange(win_rate) %>%
  mutate(top_three = rank(-win_rate, ties.method = "first") <= 3)
```

# 10. Visualisation 4: Team Win Rate

A horizontal bar chart is used because it avoids crowding long team names.

```{r fig4-team-win-rate}
fig4 <- ggplot(team_summary, aes(x = win_rate, y = reorder(team, win_rate), fill = top_three)) +
  geom_col(width = 0.7) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = c("TRUE" = col_secondary, "FALSE" = col_primary), guide = "none") +
  labs(
    title = "Some teams win more consistently than others",
    subtitle = "Overall IPL win rate by franchise",
    x = "Win rate",
    y = "Team",
    caption = "Orange highlights the top three teams by win rate among teams with at least 10 matches."
  ) +
  base_theme

fig4

ggsave("../images/fig4_team_win_rate.png", fig4, width = 10, height = 6, dpi = 300)
```

# 11. Visualisation 5: Team Scoring Heatmap

The heatmap is a higher-order graphic because it combines three dimensions: team, season, and average first innings score.

```{r fig5-heatmap}
team_season_scores <- first_innings %>%
  group_by(season_year, batting_team) %>%
  summarise(
    avg_score = mean(innings_total, na.rm = TRUE),
    matches = n(),
    .groups = "drop"
  ) %>%
  filter(matches >= 2)

fig5 <- ggplot(team_season_scores, aes(x = season_year, y = reorder(batting_team, avg_score), fill = avg_score)) +
  geom_tile(colour = "white", linewidth = 0.2) +
  scale_fill_gradient(low = "#FFF3B0", high = col_accent, name = "Avg score") +
  scale_x_continuous(breaks = sort(unique(team_season_scores$season_year))) +
  labs(
    title = "Team scoring patterns reveal adaptation over time",
    subtitle = "Average first innings score by batting team and season",
    x = "Season",
    y = "Batting team",
    caption = "Darker cells indicate higher average first innings scores."
  ) +
  base_theme +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

fig5

ggsave("../images/fig5_team_score_heatmap.png", fig5, width = 11, height = 7, dpi = 300)
```
